---
title: Streetscape Project
subtitle: CA Active Transportation Program
author: Connor P. Jackson
date: "`r format(Sys.time(), '%B %d, %Y - %H:%M %Z')`"
lang: en
output:
  ioslides_presentation:
    transition: faster
    df_print: !expr pander::pander
    css: slides.css
  html_document:
    fig_caption: yes
    number_sections: no
    highlight: textmate
    toc: no
    toc_depth: 2
    toc_float: yes
    df_print: !expr pander::pander
  pdf_document:
    fig_caption: yes
    toc: yes
    number_sections: yes
    keep_tex: yes
    pandoc_args: "--wrap=none"
    citation_package: biblatex
    md_extensions: -auto_identifiers
    df_print: !expr pander::pander
linestretch: 1.1
linkcolor: blue
urlcolor: blue
link-citations: true
bibliography: ["/Users/connor/cpjackson@berkeley.edu - Google Drive/My Drive/CitationManagement/library.bib", grateful-refs.bib]
reference-section-title: References
biblio-title: References
biblio-style: authoryear
biblatexoptions: "isbn=false, url=false"
pdfa: 2a
---

```{r setup, include=FALSE}
library(knitr)
library(pander)
library(rmarkdown)
library(grateful)
library(data.table)
library(lubridate)
# read_chunk("FILENAMEHERE")
options(knitr.kable.NA = '')
panderOptions("digits", 3)
panderOptions("table.split.cells", 15)
panderOptions("table.split.table", 120)
panderOptions("table.emphasize.rownames", FALSE)
```

```{r import-data, include = FALSE}
headers <- fread("/Users/connor/cpjackson@berkeley.edu - Google Drive/My Drive/Streetscape/ATP/headers.csv")
headers <- headers[include == TRUE, header]
proj <- fread("/Users/connor/cpjackson@berkeley.edu - Google Drive/My Drive/Streetscape/ATP/2017/ATP2017_data.csv", 
              select = headers, check.names = TRUE)

setnames(proj, gsub("form.0..", "", colnames(proj), fixed = TRUE))
setnames(proj, gsub("\\.0\\.$", "", colnames(proj)))
setnames(proj, gsub("^.._", "", colnames(proj)))
proj <- proj[PROJ_TYPE %in% c("Combination (I/NI)", "Infrastructure (I)")]

proj[, atp_year := 2017]
proj[, CONST_END_DATE := dmy(CONST_END_DATE)]
proj[, ID := paste(CT_DIST, IMP_AGCY_NAME, AGENCY_APP_NUM, sep = "-")]
```

## Active Transportation Program Overview

- Goal: use ATP approval/denial for RD identification
- ATP application forms are formulaic, so data can be easily extracted via Adobe
- Application forms for funded projects are posted online.
- Will need to request/CPRA unfunded applications
- Beginning with 2017 round of funding to explore what exists

## Machine-Readable Data

- Applying agency information
- Rough project location
  - city, county, congressional district, census block group
  - associated demographic data
- Type of project (bike, walk, safe routes to school, trail)
- Project stage actual/projected start and end dates, and costs
  - Project Approval & Environmental Documentation
  - Plans, Specifications, and Estimates
  - Right of Way acquisition
  - Construction
- Safety information and pre-application use counts
- Cost estimates, including ATP and non-ATP funds

## Machine-Readable Data

- Project component counts
  - Feet of new bike lane (by type)
  - Bike boxes, parking, bikeshare
  - signal timing and configuration changes
  - Feet of new/widened/rebuilt sidewalk
  - Physical barriers/planters
  - ADA enhancements
  - New/enhanced crosswalks, bulbouts, mid-block refuge
  - roundabouts
  - lighting, benches, trashcans, trees
  - Travel lane, turn pocket removals
  - Multi-use trail type and length

## Data still needed

- Precise location Data! 
  - Every project comes with maps, but no GIS or machine-readable spatial data
  - Exploring other public data, and talking with program administrators. Otherwise, a task for some undergraduates? 
- Project status and actual construction dates
- Project scores - published, just need to be merged

## Approved Project Components

```{r bike-ped-counts, echo = FALSE}
proj[, .N, by = .(bike_component=ST_BICYCLE_APPLIES, ped_component=ST_PED_APPLIES)]
```

```{r SRTS, echo = FALSE}
proj[, .N, by = .(SafeRoutesToSchool=ST_SRTS)]
```

## Approved Project Timelines

```{r timeline, echo = FALSE}
pander(proj[, .N, keyby = .(completion_year = year(CONST_END_DATE))], keys.as.row.names = FALSE)
```

## Left Hand Side Data

- Safegraph data acquired! 
- Traffic data more feasible than initially thought
  - TomTom/ADCi (via Matt) ($)
  - Streetlight (discounted or free)
  - INRIX (potentially free from individual cities)


